### 2016-01-07 src/3_TimeSeries.tex
saving my work

A time series is a collection of observations in time (discrete or continuous).
The analysis of time series main objective is to find possible internal
structure in the data such as autocorrelation, trend or seasonal variation.
Some of application and uses of time series analysis are data compression,
explanatory variables (relationships with other variables, seasonal factors,
etc.), signal processing, forecasting (predict future values).  There are two
main approaches to study financial time series: to study directions of financial
rates and to explain financial rate volatility.  This chapter reviews the more
relevant techniques in the rich and rapidly growing field of time series
analysis considering these two approaches.

\section{Characteristics of Time Series}

@@

\section{Stylized facts of asset returns}
\label{sec:stylizedfacts}

Returns of a time series $y$ is defined as:
\begin{equation}
\label{eq:returndef}
r_t = y_t - y_{t-1}  \, ,
\end{equation}

\begin{description}
\item[Dependence] autocorrelation in returns is largely insignificant. However,
autocorrelation in the absolute and squared returns is always positive and
significant and decays slowly. In addition, the autocorrelation in the absolute
returns is generally higher than the autocorrelation in the corresponding
squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
 \ref{fig:returnacf}. The figure shows how the correlations are significant even
 for very long lags, this imply that a long-memory process because of its
 long-range dependence. 
 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. Figure
\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
in average Monday has negative returns. However, this is changing in time,
figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
three months and it shows that Mondays and Wednesdays has positive returns in
average.

@@

\section{Vector Autoregressive Models}\label{sec:varvec}

>>>>

A time series is a collection of observations in time (discrete or continuous).
The analysis of time series main objective is to find possible internal
structure in the data such as autocorrelation, trend or seasonal variation.
Some of application and uses of time series analysis are data compression,
explanatory variables (relationships with other variables, seasonal factors,
etc.), signal processing and forecasting (predict future values), which is the focus of this thesis. 
There are two main approaches to study financial time series: to study directions of financial
rates and to explain financial rate volatility.  This chapter reviews the most
relevant techniques in the rich and rapidly growing field of time series
analysis considering these two approaches.

\section{Characteristics of Financial Time Series}
%\section{Stylized facts of asset returns}
\label{sec:stylizedfacts}

Returns of a time series $y_t$ is defined as:
\begin{equation}
\label{eq:returndef}
r_t = y_t - y_{t-1}  \, ,
\end{equation}

\begin{description}
\item[Dependence] autocorrelation function (ACF) in returns is largely insignificant. The ACF measures the linear predictability of the series at time $t$ using only the value at time $s$:

\begin{equation}
\rho(s,t) = \frac{\gamma(s,t)}{\sqrt{\gamma(s,s)\gamma(t,t)}}
\end{equation}
\noindent where $\gamma(s,t)$ is the autocovariance function measures the linear dependence between two points on the same series observed at different times, it is defined as the second moment product:
\begin{equation}
\gamma_y(s,t) = E[(y_s-\mu_s)(y_t - \mu_t)]
\end{equation}
However, the autocorrelation in the absolute and squared returns is always positive,
significant and decays slowly. In addition, the autocorrelation in the absolute
returns is generally higher than the autocorrelation in the corresponding
squared returns. The returns autocorrelation function (ACF) of SPY is shown in figure
 \ref{fig:returnacf}. The figure shows how the correlations are significant even
 for very long lags, this imply that a long-memory process because of its
 long-range dependence. 
 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. Figure
\ref{fig:mondayeffect} shows the Monday effect from 1928 to 2007. It shows that
in average Monday has negative returns. However, this is changing in time,
figure \ref{fig:mondayeffect2} shows returns only for year 2007 and its last
three months and it shows that Mondays and Wednesdays has positive returns in
average.

@@

@@

%\section{Vector Autoregressive Models}\label{sec:varvec}
\section{Multivariate Time Series modelling}
